Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-04-09 manuscript/paper.tex
readthrough for typos; incorporated Ana's explanation for Fig 2 caption

Visualisation is fundamentally important to data analysis.
Many specialised tools exist to aid the visual analysis and 
quality control
of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise 
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the 
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

@@

\subsection{Data model} \label{subsec:Data_Model} 
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
ARGS~\citep{wong2024general}. This efficient ARG encoding 
is implemented by the 
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,korfmann2023weak,
tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about the 
nodes and edges that describe the graph topology and the sites 
and mutations encoding the sequence variation (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
% The Node and Edge tables describe the genealogical relationships
% between the sampled and ancestral sequences, 
% while the Mutation table describes
% sequence variation contained in the samples. 
% (see \hyperref[subsec:User_Interface]{User Interface}).

@@

\subsection{Applications}
The purpose of \texttt{tsbrowse} is to provide an interactive view of the 
\texttt{tskit} ARG data model to guide intuition, improve inference quality control
and facilitate debugging. An example of how we can deepen our understanding
using the genome-browser-like perspective provided by \texttt{tsbrowse}
is provided the simulated ARG of Fig~\ref{fig:Figure_2}, where 
the gap in the edges view corresponds to the characteristic dip in diversity
of a selective sweep. Comparing this ground truth to the ARGs inferred
by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
we can see that there are substantial qualitative differences between the
results. These differences in ancestral haplotypes illustrated by 
the edges view are unlikely to be captured by the tree-by-tree distance
metrics usually used to evaluate
inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
providing further motivation for new and improved ways to compare
simulated and inferred ARGs~\citep{fritze2024forest}.

>>>>

Visualisation is fundamentally important to data analysis.
Many specialised tools exist to aid the visual analysis and
quality control
of genome assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

@@

\subsection{Data model} \label{subsec:Data_Model}
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of
ARGS~\citep{wong2024general}. This efficient ARG encoding
is implemented by the
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
    haller2019tree,baumdicker2022efficient,
    adrion2020community,lauterbur2023expanding,korfmann2023weak,
    tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
    mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about the
nodes and edges that describe the graph topology and the sites
and mutations that encode the sequence variation (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
% The Node and Edge tables describe the genealogical relationships
% between the sampled and ancestral sequences, 
% while the Mutation table describes
% sequence variation contained in the samples. 
% (see \hyperref[subsec:User_Interface]{User Interface}).

@@

\subsection{Applications}
The purpose of \texttt{tsbrowse} is to provide an interactive view of the
\texttt{tskit} ARG data model to guide intuition, improve inference quality
control and facilitate debugging. An example of how we can deepen our understanding
using the genome-browser-like perspective provided by \texttt{tsbrowse}
is given in the simulated ARG of Fig~\ref{fig:Figure_2}, where
the gap in the edges view corresponds to the characteristic dip in diversity
of a selective sweep. Comparing this ground truth to the ARGs inferred
by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
we can see that there are substantial qualitative differences between the
results. These differences in ancestral haplotypes illustrated by
the edges view are unlikely to be captured by the tree-by-tree distance
metrics usually used to evaluate
inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
providing further motivation for new and improved ways to compare
simulated and inferred ARGs~\citep{fritze2024forest}.
